---
title: "Summary of Microbe Associated Compounds"
date: "`r Sys.Date()`"
params:
  node_file: "default_node.csv"
  edge_file: "default_edge.csv"
---

```{css, echo=FALSE}
h1, h2, h3, h4 {
  text-align: center;
}
```

```{r, setup, echo = F}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = T, echo = F, 
                      fig.align = "center")
options(scipen = 999)
```

```{r}
# load necessary packages 
# Set CRAN mirror for non-interactive installation
options(repos = c(CRAN = "https://cloud.r-project.org"))

packages <- c("dplyr", "ggplot2", "readr", "cowplot", "thematic", "tidyr",
              "tibble", "plotly", "rmarkdown", "ggpubr", "stringr", "forcats")

for (package in packages) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package)
    library(package, character.only = TRUE)
  }
}
```

```{r}
# make a function to calculte jaccard similarity between food items
jaccard_similarity <- function(food1, food2){ 
  intersection <- length(intersect(food1, food2))
  union <- length(food1) + length(food2) - intersection
  return(intersection/union)
}
```

```{r}
# get data 
hmdb <- read_csv("../Data/hmdb.csv")
nodes <- read_csv(params$node_file)
edges <- read_csv(params$edge_file)

# clean node file 
microbe_comps <- nodes %>% 
  filter(origin != 'food' & origin != 'none') %>% 
  mutate(assoc_food = gsub("\\[|\\]|'", "", assoc_food)) %>%  # remove [], '
  separate_rows(assoc_food, sep = ",")  # split into multiple rows
microbe_comps$assoc_food <- stringr::str_trim(microbe_comps$assoc_food)

# subset edge columns 
edges_subset <- edges %>% 
  select(compound2, organisms)
colnames(edges_subset)[1] <- "compound"

# merge node and edge data 
microbe_comps <- microbe_comps %>% 
  left_join(edges_subset, by = "compound", relationship = "many-to-many") %>% 
  mutate(organisms = gsub("\\[|\\]|'", "", organisms)) %>%  # remove [], '
  separate_rows(organisms, sep = ",")  # split into multiple rows
microbe_comps$organisms <- stringr::str_trim(microbe_comps$organisms)

# merge hmdb information 
colnames(hmdb)[1] <- "compound"
meta <- microbe_comps %>% 
  left_join(hmdb, by = "compound", relationship = "many-to-many")
```

Here are visuals to aid in understanding compounds that microbes are predicted to have created. These compounds are located in the nodes and edges file that are generated in the graph creation portion of the pipeline. Compounds included are those with origin "microbe" or "both", the latter being both microbe or food originating. This report is broken up into three sections: (1) understanding compounds being produced by a genus; (2) understanding compounds being produced by species; (3) understanding compounds that can come from both microbes and food and the genus and species level.

## Production of Compounds by Genera {.tabset}

### Distinct Compounds 

<center>
```{r}
# separate genus and species 
taxonomy_meta <- meta %>% 
  select(compound, organisms, taxonomy_super_class, taxonomy_class) %>% 
  distinct(compound, organisms, .keep_all = T) %>% #there could be repetes due to food items
  separate_wider_delim(organisms, delim = '.', names = c('genus', 'species'), 
                       too_few = "align_start", too_many = "merge", cols_remove = F)
taxonomy_meta %>% 
  group_by(genus) %>% 
  summarise(n_compounds = n_distinct(compound)) %>% 
  mutate(genus = fct_reorder(genus, n_compounds)) %>% 
  ggplot(aes(x = genus, y = n_compounds)) + 
    geom_bar(stat = "identity") + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
    xlab("") +
    ylab("Number Unique of Compounds") -> p 

ggplotly(p)
```
</center>

### Compound Taxonomy 

<center>
```{r}
taxonomy_meta %>% 
  count(genus, taxonomy_super_class) %>% 
  ggplot(aes(y=taxonomy_super_class, x=genus)) +
    geom_point(aes(color = n), size = 1) + 
    scale_color_gradient(low = "lightgrey", high = "purple4") +
    theme_bw() + 
    theme(axis.text.x = element_blank()) + 
    xlab("Genus") + 
    ylab("Taxonomy Super Class ") -> p 

ggplotly(p)
```

```{r}
taxonomy_meta %>% 
  count(genus, taxonomy_class) %>% 
  ggplot(aes(y=taxonomy_class, x=genus)) +
    geom_point(aes(color = n), size = 1) + 
    scale_color_gradient(low = "lightgrey", high = "purple4") +
    theme_bw() + 
    theme(axis.text.x = element_blank()) + 
    xlab("Genus") + 
    ylab("Taxonomy Class ") -> p 

ggplotly(p)
```
</center>

### Similarity Metrics 

<center>
```{r}
# must remove NAs for this portion 
jaccard_df <- taxonomy_meta %>% 
  select(compound, genus) %>% 
  na.omit()

# get list of food items
genera <- unique(jaccard_df$genus)

# precompute compound lists for each genus (for speed)
genus_compounds <- lapply(genera, function(g) {
  unique(jaccard_df$compound[jaccard_df$genus == g])
})
names(genus_compounds) <- genera

# initialize empty matrix
n <- length(genera)
jaccard_matrix <- matrix(NA, nrow = n, ncol = n,
                         dimnames = list(genera, genera))

# compute only lower triangle
for (i in seq_along(genera)) {
  for (j in seq_len(i)) {  # compute only for j ≤ i
    genus1 <- genera[i]
    genus2 <- genera[j]
    
    score <- jaccard_similarity(genus_compounds[[genus1]], genus_compounds[[genus2]])
    
    jaccard_matrix[i, j] <- score
    jaccard_matrix[j, i] <- score  # mirror to upper triangle
  }
}

# create a long format for matrix
jaccard_long <- as.data.frame(jaccard_matrix) %>% 
  rownames_to_column(var = "Row") %>% 
  pivot_longer(cols = -Row, names_to = "Column", values_to = "Value")

# visualize matrix
ggplot(jaccard_long, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() + 
  scale_fill_gradient(low = "lightblue", high = "darkblue") + 
  labs(title = "Jaccard Matrix Heatmap", x = "Genus", y = "Genus", fill = "Simiarity Score") +
  theme_bw() + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        plot.title = element_text(hjust = 0.5)) -> p

plotly::ggplotly(p)
```
</center>

## Production of Compounds by Species {.tabset}

### Distinct Compounds 

<center>
```{r}
taxonomy_meta %>% 
  group_by(species) %>% 
  summarise(NumberOfCompounds = n_distinct(compound)) %>% 
  mutate(species = fct_reorder(species, NumberOfCompounds)) -> p_df 
  
ggplot(p_df, aes(x = species, y = NumberOfCompounds)) + 
  geom_bar(stat = "identity") + 
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("Species") +
  ylab("Number of Compounds") + 
  annotate("text", 
         x = length(p_df$species)/4, 
         y = max(p_df$NumberOfCompounds)/2 , 
         label = paste0("Numer of Species: ", length(p_df$species),
                        "\nMean \u00B1 SD: ", round(mean(p_df$NumberOfCompounds), 2), 
                            "\u00B1", round(sd(p_df$NumberOfCompounds), 2), 
                        "\nMedian: ", round(median(p_df$NumberOfCompounds), 2), 
                        "\nRange: ", round(range(p_df$NumberOfCompounds)[1], 2), "-", 
                            round(range(p_df$NumberOfCompounds)[2], 2)
                        )
         )-> p 

ggplotly(p)
```
</center>

### Top 10% of Producers

<center>
```{r}
perc10 <- round(length(p_df$species)*0.1, 0) # get a number for 10%
p_df_filter <- head(p_df, perc10)            # filter for top 10%
perc10species<- p_df_filter$species          # get list of top 10% species

taxonomy_meta %>% 
  dplyr::filter(species %in% perc10species) %>% 
  count(species, taxonomy_super_class) %>% 
  ggplot(aes(y=taxonomy_super_class, x=species)) +
    geom_point(aes(color = n), size = 1) + 
    scale_color_gradient(low = "lightgrey", high = "purple4") +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
    xlab("") + 
    ylab("Taxonomy Super Class ") -> p1

taxonomy_meta %>% 
  dplyr::filter(species %in% perc10species) %>% 
  count(species, taxonomy_class) %>% 
  ggplot(aes(y=taxonomy_class, x=species)) +
    geom_point(aes(color = n), size = 1) + 
    scale_color_gradient(low = "lightgrey", high = "purple4") +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
    xlab("") + 
    ylab("Taxonomy Class ") -> p2

ggplotly(p1)
ggplotly(p2)
```
</center>

## Identifying Compounds Both Microbes and Food Produce {.tabset}

```{r}
tax_food_meta <- meta %>% 
  select(compound, organisms, assoc_food, taxonomy_super_class, taxonomy_class) %>% 
  distinct(compound, organisms, assoc_food, .keep_all = T) %>% 
  separate_wider_delim(organisms, delim = '.', names = c('genus', 'species'), 
                       too_few = "align_start", too_many = "merge", cols_remove = F)
```

### Genera x Food Items 

<center>
```{r, fig.height=20}
tax_food_meta %>% 
  group_by(genus, assoc_food) %>% 
  summarise(n_compound = n_distinct(compound)) %>% 
  ggplot(aes(x = assoc_food, y = genus, fill = n_compound)) + 
    geom_tile() + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
    xlab("Food Item") + 
    ylab("Genus") -> p 

ggplotly(p)
```
</center>

### Species x Food Items
 
 <center>
```{r, fig.height=60}
tax_food_meta %>% 
  group_by(species, assoc_food) %>% 
  summarise(n_compound = n_distinct(compound)) %>% 
  ggplot(aes(x = assoc_food, y = species, fill = n_compound)) + 
    geom_tile() + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
    xlab("Food Item") + 
    ylab("Species") -> p 

ggplotly(p)
```
</center>