---
title: "Metabolome Comparison Report"
date: "`r Sys.Date()`"
params:
  patterns: "default.csv"
  metabolome: "metab_default.csv"
---

```{css, echo=FALSE, purl=FALSE}
h1, h2, h3, h4 {
  text-align: center;
}
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = TRUE, echo = FALSE,
                      fig.align = "center")
options(scipen = 999)

library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(plotly)
```

```{r load-data}
hmdb <- read_csv("../Data/hmdb.csv")
colnames(hmdb)[1] <- "compound"
hmdb_subset <- hmdb %>%
  dplyr::select(compound, taxonomy_super_class, taxonomy_class)

patterns   <- read_csv(params$patterns)
metabolome <- read_csv(params$metabolome)

# ── Detect whether this is a 1-edge or 2-edge CSV ───────────────────────────
# 1-edge CSVs have compound1_id / compound2_id
# 2-edge CSVs additionally have compound3_id / edge2_* columns
has_two_edges <- "compound3_id" %in% colnames(patterns)
```

```{r helper-functions}
# Collect all compound id+origin pairs present in a pattern subset.
# Works for both 1-edge and 2-edge CSVs.
get_compounds <- function(df) {
  pairs <- list(
    df %>%
      dplyr::select(compound = compound1_id, origin = compound1_origin)
  )

  if ("compound2_id" %in% colnames(df)) {
    pairs[[2]] <- df %>%
      dplyr::select(compound = compound2_id, origin = compound2_origin)
  }

  if (has_two_edges && "compound3_id" %in% colnames(df)) {
    pairs[[3]] <- df %>%
      dplyr::select(compound = compound3_id, origin = compound3_origin)
  }

  bind_rows(pairs) %>% distinct()
}

# Filter to metabolome-observed compounds, join taxonomy, and return.
enrich_with_metabolome <- function(df) {
  df %>%
    filter(compound %in% metabolome$compound) %>%
    left_join(hmdb_subset, by = "compound")
}

# Dot-matrix plots of compound count by origin × taxonomy level.
plot_taxonomy_dots <- function(comp_orig) {
  for (tax_level in c("taxonomy_super_class", "taxonomy_class")) {
    label <- gsub("_", " ", str_to_title(tax_level))
    p <- comp_orig %>%
      count(origin, .data[[tax_level]]) %>%
      ggplot(aes(y = .data[[tax_level]], x = origin)) +
        geom_point(aes(color = n), size = 1) +
        scale_color_gradient(low = "lightgrey", high = "purple4") +
        theme_bw() +
        xlab("Origin") +
        ylab(label) +
        labs(color = "Count")
    print(p)
  }
}

# Full tabset section for one pattern subset.
# section_label  : displayed in the heading, e.g. "Diet → Microbe → Host"
# description    : paragraph shown above the tabset
render_section <- function(pattern_subset, section_label, description) {
  comp_orig <- get_compounds(pattern_subset) %>%
    enrich_with_metabolome()

  n_shared <- n_distinct(comp_orig$compound)

  cat(sprintf("\n\n## %s\n\n", section_label))
  cat(description, "\n\n")

  if (n_shared == 0) {
    cat(
      "> **No compounds from this pattern set were detected in the metabolome.**\n\n"
    )
    return(invisible(NULL))
  }

  cat(sprintf(
    "*%d unique compound(s) from this pattern set were also detected in the metabolome.*\n\n",
    n_shared
  ))

  cat("## {.tabset}\n\n")

  # Tab 1: taxonomy summary
  cat("### Compound Taxonomy Summary\n\n")
  cat("```{r, echo=FALSE}\n")  # placeholder — actual plots rendered below
  cat("```\n\n")
  plot_taxonomy_dots(comp_orig)

  # Tab 2: compound list
  cat("\n\n### List of Shared Compounds\n\n")
  print(
    comp_orig %>%
      dplyr::select(compound, origin) %>%
      distinct() %>%
      arrange(origin, compound),
    n = Inf
  )
}
```

This report cross-references compounds found in graph query pattern results against an observed metabolome. Only compounds present in both datasets are shown. Taxonomy annotations are drawn from the HMDB reference.

```{r split-patterns}
# ── Most restrictive ─────────────────────────────────────────────────────────
# 1-edge:  compound1 origin = diet  AND  compound2 origin = microbe
# 2-edge:  additionally compound3 origin = host
if (has_two_edges) {
  patterns_strict <- patterns %>%
    filter(compound1_origin == "diet" &
           compound2_origin == "microbe" &
           compound3_origin == "host")
} else {
  patterns_strict <- patterns %>%
    filter(compound1_origin == "food" &
           compound2_origin == "microbe")
}

# ── Least restrictive ────────────────────────────────────────────────────────
# Everything that does NOT satisfy the strict criteria above
patterns_loose <- patterns %>%
  anti_join(patterns_strict,
            by = intersect(colnames(patterns_strict), colnames(patterns)))
```

---

## Most Restrictive Results: `r if (has_two_edges) "Diet → Microbe → Host" else "Food → Microbe"`

`r if (has_two_edges) "Patterns where the starting compound is exclusively dietary, the intermediate compound is exclusively microbial, and the endpoint compound is exclusively host-derived — the most resolved diet-to-microbe-to-host metabolic chain." else "Patterns where the starting compound is exclusively from food and the product compound exclusively originates from microbes."`

## {.tabset}

### Compound Taxonomy Summary

```{r strict-taxonomy}
comp_strict <- get_compounds(patterns_strict) %>% enrich_with_metabolome()

if (nrow(comp_strict) == 0) {
  cat("> **No compounds from this pattern set were detected in the metabolome.**\n")
} else {
  plot_taxonomy_dots(comp_strict)
}
```

### List of Shared Compounds

```{r strict-list}
if (nrow(comp_strict) > 0) {
  print(
    comp_strict %>%
      dplyr::select(compound, origin) %>%
      distinct() %>%
      arrange(origin, compound),
    n = Inf
  )
}
```

---

## Least Restrictive Results: All Other Patterns

All pattern combinations that do not satisfy the most restrictive criteria above. This aggregates chains where any node has a mixed, shared, or alternative origin (e.g. `microbediet`, `hostmicrobe`, `all`, `both`).

## {.tabset}

### Compound Taxonomy Summary

```{r loose-taxonomy}
comp_loose <- get_compounds(patterns_loose) %>% enrich_with_metabolome()

if (nrow(comp_loose) == 0) {
  cat("> **No compounds from this pattern set were detected in the metabolome.**\n")
} else {
  plot_taxonomy_dots(comp_loose)
}
```

### List of Shared Compounds

```{r loose-list}
if (nrow(comp_loose) > 0) {
  print(
    comp_loose %>%
      dplyr::select(compound, origin) %>%
      distinct() %>%
      arrange(origin, compound),
    n = Inf
  )
}
```

---

## Combined Overview

Summary across both pattern sets showing how many metabolome-observed compounds each captures, and their overlap.

```{r combined-summary}
strict_ids <- comp_strict %>% pull(compound) %>% unique()
loose_ids  <- comp_loose  %>% pull(compound) %>% unique()

both_ids   <- intersect(strict_ids, loose_ids)
only_strict <- setdiff(strict_ids, loose_ids)
only_loose  <- setdiff(loose_ids,  strict_ids)

summary_df <- tibble(
  Category = c(
    "Most restrictive only",
    "Least restrictive only",
    "In both pattern sets",
    "Total unique"
  ),
  `Compound count` = c(
    length(only_strict),
    length(only_loose),
    length(both_ids),
    length(union(strict_ids, loose_ids))
  )
)

knitr::kable(summary_df, align = "lr")
```

```{r venn-bar}
# Stacked bar showing compound set membership by origin
bind_rows(
  comp_strict %>% mutate(pattern_set = "Most restrictive"),
  comp_loose  %>% mutate(pattern_set = "Least restrictive")
) %>%
  distinct(compound, origin, pattern_set) %>%
  count(pattern_set, origin) %>%
  ggplot(aes(x = pattern_set, y = n, fill = origin)) +
    geom_bar(stat = "identity", position = "stack") +
    theme_bw() +
    xlab("Pattern Set") +
    ylab("Number of Compounds") +
    labs(fill = "Origin") +
    scale_fill_brewer(palette = "Set2")
```
